---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CohortConstructor

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/CohortConstructor)](https://CRAN.R-project.org/package=CohortConstructor)
[![codecov.io](https://codecov.io/github/oxford-pharmacoepi/CohortConstructor/coverage.svg?branch=main)](https://app.codecov.io/github/oxford-pharmacoepi/CohortConstructor?branch=main)
[![R-CMD-check](https://github.com/oxford-pharmacoepi/CohortConstructor/workflows/R-CMD-check/badge.svg)](https://github.com/oxford-pharmacoepi/CohortConstructor/actions)
[![Lifecycle:Experimental](https://img.shields.io/badge/Lifecycle-Experimental-339999)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of CohortConstructor is to help on the creation and manipulation of cohorts in the OMOP Common Data Model.

## Installation

You can install the development version of CohortConstructor from [GitHub](https://github.com/) with:

``` {r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("oxford-pharmacoepi/CohortConstructor")
```

## Example usage

``` {r}
library(CDMConnector)
library(PatientProfiles)
library(CohortConstructor)

con <- DBI::dbConnect(duckdb::duckdb(), dbdir = eunomia_dir())
cdm <- cdm_from_con(con, cdm_schema = "main", 
                    write_schema = c(prefix = "my_study_", schema = "main"))
```

### Generating concept based cohorts
``` {r}
cdm <- generate_concept_cohort_set(cdm = cdm, 
                            name = "medications",
                            concept_set = list("diclofenac" = 1124300,
                                               "acetaminophen" = 1127433))
cohort_count(cdm$medications)
cohort_attrition(cdm$medications)
```

### Applying demographic requirements
``` {r}
cdm$medications %>% 
  requireDemographics(ageRange = list(c(40, 65)),
                      sex = "Female")
cohort_count(cdm$medications)
cohort_attrition(cdm$medications)
```

### Require presence in another cohort
We can also require that individuals are in another cohort over some window. Here for example we require that study participants are in a GI bleed cohort any time prior up to their entry in the medications cohort.
```{r}
cdm <- generate_concept_cohort_set(cdm = cdm, 
                            name = "gibleed",
                            concept_set = list("gibleed" = 192671))

cdm$medications <- cdm$medications %>% 
  requireCohortIntersectFlag(targetCohortTable = "gibleed",
                             window = c(-Inf, 0))

cohort_count(cdm$medications)
cohort_attrition(cdm$medications)
```


### Combining cohorts

Generate a combination cohort.

```{r}
cdm <- generateCombinationCohortSet(cdm = cdm, 
                                    name = "combinations", 
                                    targetCohortName = "medications")

cohortSet(cdm$combinations)
cohortCount(cdm$combinations)


cdmDisconnect(cdm)
```

